
---
title: "Reproducible Research - Project 1"
author: "NK"
date: "Tuesday, May 12, 2015"
output: html_document
---
The following document is a code for Project 1 of reproducible research. It describes an exploratory data analysis on a step counter data. 

First set global options to always display code chunks:
```{r setoptions, echo=TRUE}
library(knitr)
opts_chunk$set(echo=TRUE)
```



##Loading and Preprocessing the Data

First, set the working directory
```{r}
#Set working directory
setwd("C:\\Users\\N\\Dropbox\\Coursera\\05 - Reproducible Research\\RepData_PeerAssessment1")

```

Read accelerometer data that is located in the same folder
```{r, echo=TRUE}
#read in data
dat=read.csv(file="activity.csv",
             header=TRUE, sep=",",
             stringsAsFactors=FALSE,
             na.strings="NA")
dat$date=as.factor(dat$date)
```

##What is the average dailty activity pattern?

**Calculate the total number of steps taken per day**

First extract the step data and the sum the steps via the aggregate function.
```{r, echo=TRUE}
#Take the step data and remove the NA values
mean_step_data=dat[!is.na(dat$steps),1:3]
#Calculate the sum
sum_steps=aggregate(mean_step_data[1],
                    by=list(mean_step_data$date),
                    FUN=sum)
sum_steps
```

**Create a histogram of the step summaries**

Create and show a histogram plot:
```{r, echo=TRUE}
#Make historgram 
hist(sum_steps$steps,
     breaks=10,
     xlab="Daily Steps",
     col="red")
```

**Find the mean and median**

The median of the number of steps is:
```{r, echo=TRUE}
median(sum_steps$steps)
```
The mean number of steps is:
```{r, echo=TRUE}
mean(sum_steps$steps)
```

---
##What is the average daily activity pattern?
**Make a time series plot:**

Find the average interval number of steps:
```{r, echo=TRUE}
#interval data mean
id=aggregate(mean_step_data[1],by=list(mean_step_data$interval),FUN=mean)
```

Plot the mean interval data:
```{r, echo=TRUE}
plot(unique(mean_step_data$interval),
     id$steps,pch=16,
     type="l",
     xlab="Daily Interval",
     ylab="Average # of steps"
     )
```

**Find the Interval with maximum steps**

The values are stored in the id variable. Find the maximum index:
```{r,echo=TRUE}
which.max(id$steps)
```
with a maximum value of
```{r,echo=TRUE}
max(id$steps)
```

##Imputing missing Values
**Find total missing entries**

First create a vector full of missing steps entries and then report its length
```{r,echo=TRUE}
#create empty vector indicator
na_vec=is.na(dat$steps)
#find total entires that are missing
sum(na_vec)
```

**Fill missing entries**

There are two strategies to fully get rid of NA values:
1. First replace them by the mean of the day

The code first appends a column full of means. The code then replaces the missing values with those means.

```{r}
#append means
dat$means<-ave(dat$steps,dat$date,rm.na=TRUE)
#remove NAs
dat[na_vec,"steps"]=dat[na_vec,"means"]
```

**Assign to new Tidy Data set**

label the tidy data set as tdat
```{r,echo=TRUE}
#find values that are still NA
still_na=is.na(dat$means)
#Assign tidy data set
tdat=dat[!still_na,]
#verify that there are no empties:
sum(is.na(tdat))
```

**Make a histogram of the total number of steps**

Calculate sum and draw histogram:
```{r,echo=TRUE}
#Calculate the sum
sum_step2=aggregate(tdat[1],
                    by=list(tdat$date),
                    FUN=sum)
#Make histogram 
hist(sum_step2$steps,
     breaks=10,
     xlab="Daily Steps",
     col="red")
```
The median of the tidy number of steps is:
```{r, echo=TRUE}
median(sum_step2$steps)
```
The mean tudy number of steps is:
```{r, echo=TRUE}
mean(sum_step2$steps)
```
The mean and median has not changed by replacing NA values with the mean step values.

##Difference in Activity Patterns between Weekends and Weekdays
**Separate data into wekedays**

The following code first appends a weekdays variable.
Then the code appends a variable that states whether the day is a weekend or a weekday.
```{r}
#change factors to characters
tdat$date<-lapply(tdat[,"date"],as.character)
#change char to date
library("lubridate")
tdat$date<-ymd(tdat$date)
tdat$weekdays<-weekdays(tdat$date)
tdat$isWeekend<-tdat$weekdays %in% c('Saturday','Sunday')
```

**Determine patterns in Intervals and Weekdays**

First prepare the interval data (id) to properly rename the variables and calculate the means
```{r}
#Calculate means and rename variables and prepare values for plotting
id=aggregate(tdat[,1],by=list(tdat$interval,tdat$isWeekend),FUN=mean)
#rename variables
library(plyr)
id=rename(id,c("Group.1"="interval", "Group.2"="isWeekend","x"="steps"))
id$isWeekend<-as.factor(ifelse(id$isWeekend, "weekend", "weekday"))
```

Now run some plotting code to reproduce the graph usign the lattice plot function:

```{r}
#Plot 
library(lattice)
xyplot(steps~interval| isWeekend, data=id,type="l")


```

There is a clear difference in the shapes of the weekeday steps and the weekend step patterns. There are more steps in the weekend and the peak of the weekdays occurs a early in the interval.






